Retrofitting Effect Handlers onto OCaml

### 2020-10-28 retro-concurrency.tex
Section 2

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			%keywords=[3]{int,string}, %types%
			%keywordstyle=[3]\color{caribbeangreen},
			keywords=[4]{effect,perform,continue,discontinue},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with fast
foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables to obtain a backtrace. Our primary motivation for adding effect handlers
to OCaml is to support (1) highly concurrent applications such as web servers to
be expressed naturally in \emph{direct-style} as opposed to using callbacks and
(2) efficient data and task parallel programs in the vein of OpenMP~\cite{} and
Intel TBB~\cite{}.

@@

>>>>

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			keywords=[3]{effect,perform,continue,discontinue},
			keywordstyle=[3]\color{caribbeangreen}\bfseries,
			keywords=[4]{call,return,external},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables to obtain a backtrace. Our primary motivation for adding effect handlers
to OCaml is to support (1) highly concurrent applications such as web servers
to be expressed naturally in \emph{direct-style} as opposed to using callbacks
and (2) efficient data and task parallel programs in the vein of Cilk~\cite{},
OpenMP~\cite{} and Intel TBB~\cite{}.

@@

\begin{figure*}
\begin{minipage}{0.59\linewidth}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}
external ocaml_to_c : unit -> int = "ocaml_to_c"
exception E1
exception E2

let c_to_ocaml () = raise E1
let _ = Callback.register "c_to_ocaml" c_to_ocaml

let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c () with E2 -> 0
  with E1 -> 42;;

let _ = assert (omain () = 42)
		\end{lstlisting}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
	\end{minipage}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}[language=c]
#include <caml/mlvalues.h>
#include <caml/callback.h>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
		\end{lstlisting}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
	\end{minipage}
\end{minipage}
  \begin{minipage}{0.39\linewidth}
    \centering
    \includegraphics[scale=0.5]{figures/stock_stack}
    \subcaption{Stack layout before raising exception.}
    \label{fig:stock_stack}
  \end{minipage}
\end{figure*}

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. To this end, in this section, we provide an
overview of the program stack and related mechanisms for stock OCaml.

In order to illustrate the layout of the stock OCaml stack, consider the
program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
main function |omain| installs two exception handlers |h1| and |h2| to handle
exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
which in turn calls back into OCaml function |c_to_ocaml|, which raises the
exception |E1|. OCaml supports raising exceptions in C as well as throwing
exceptions across external calls. Hence, the exception |E1| gets caught in the
handler |h1|, and |omain| returns |42|. The layout of the stack in native code
backend just before raising the exception in |c_to_ocaml| is illustrated in
Figure~\ref{fig:stock_stack}. The stack grows downwards.

OCaml uses the hardware stack as C, and hence the program stack has alternating
sequences of C and OCaml frames. Unlike C, OCaml does not create pointers into
OCaml frames. OCaml uses the hardware support for |call| and |return|
instructions for function calls and returns. OCaml does not perform explicit
stack overflow checks in code, and, just like C, relies on the guard page at
the end of the stack region to detect stack overflow. Stack overflow is
detected by a memory fault, and |Stack_overflow| exception is raised to unwind
the stack.

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |amd64| backend, a register each is
pinned for the \emph{allocation pointer} into the minor heap used for bump
pointer allocation, and the |Caml_state|, a table of global variables used by
the runtime. These registers are chosen to be C callee saved registers. This
makes external calls extremely fast in OCaml. If the external function does not
touch the OCaml heap, then the external functions can be called directly, and
no bookkeeping is necessary. For external functions that touch the OCaml heap,
the cached allocation pointer is saved to a field in the |Caml_state|, and
loaded back on return. Similarly, callbacks into OCaml are also cheap.
Callbacks into OCaml involves loading the arguments in the right registers and
calling the OCaml function. OCaml callbacks are relatively common as
finalisation functions are executed as OCaml callbacks.

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (pc)
of the handler and the current exception pointer (a field in |Caml_state|). The
exception pointer is also updated to be the current stack pointer (sp). This
creates a linked-list of exception handler frames on the stack as shown in
Figure~\ref{fig:stock_stack}. The saved exception pointer is restored when an
exception is handled or the handled computation returns normally with a value.

In order to forward exceptions across C frames, the C stub function
|caml_start_program|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception the innermost OCaml exception handler (|raised_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_error|) if there
are no enclosing handlers. Exceptions are so cheap in OCaml that it is very
common to use them for \emph{local} control flow.

\kc{gdb backtrace}
